Asymmetric tail association in extreme climatic events affects population-extinction risk

### 2019-10-15 MainText.Rmd

<!--test hypothesis from Q2 : ext risk with density-dependence metapopulation model with r from different regime-->
To test our hypothesis from (Q2) of \nameref{Introduction} for each metapopulation model,
we estimated extinction risk for a wide range of parameter $r$ 
that covered the undercompensatory and overcompensatory regime. We 
chose equispaced values of $r$ from the 95\% quantiles 
of possible range ($r_{min},r_{bf}$), where $r_{min}$ was the minimum value of $r$ 
to get a positive, finite equilibrium $P_e$ (for details, see Appendix 
\ref{SM-model_stability}) to start with (at $t=0$). For each chosen $r$,
we calculated extinction risk after $25$ years with each metapopulation model 
where population could disperse locally (or globally) with varying dispersal 
rate $d= 0,0.1,0.2,..0.5$ and moderate to extreme left 
tail- and right-tail associated environmental noises with same Kendall's $\tau$ (= `r ktau_common`)
in different simulations (for code see [`mydispersal.R`](https://github.com/sghosh89/ERC/blob/master/mydispersal.R)).

<!--start from here-->

<!--test hypothesis from Q3 : switching of ext risk with density-dependence metapopulation model with d=0-->
Given hypothesis from (Q2) was valid - that left tail dependent noise made extinction risk 
higher in undercompensatory regime ($r < r_c$) and
right tail dependent noise made extinction risk higher
in overcompensatory regime ($r > r_c$) - we tested our third hypothesis for (Q3).
We expected the switching to take place near $r_c$, not exactly at $r = r_c$ because the transition point $r_c$
from under to overcompensatory dynamics differed in stochastic environment than its 
deterministic analog depending on the underlying nonlinearity of the model.
We plotted extinction risk along y-axis
for both left tail and right tail dependent noises 
against the varying range of $r$ along x-axis when no dispersal ($d=0$) was considered.
We excluded the possibility of dispersal in this case to avoid any rescue effect among patches. 
For each model with extreme (or moderate) left- and right-tail dependent noises, we calculated 
2.5$^{th}$ and 97.5$^{th}$ quantiles
of 10000 simulations with the probability of computed
extinction risk at each $r$. We compared $r_c$ and the value of $r$ at which 
95\% CI of extinction risk from both noises overlapped
with each other. This could give us an idea for supporting our hypothesis.

@@

For single patch linear age-structured models, we found higher extinction risk for moderate left-tail 
association (modeled by Clayton copula) in fluctuating vital 
rates compared with the one when vital rates were right tail dependent (modeled by Survival Clayton copula).
It was expected because an extreme "bad" year (e.g. with severe cold) can suppress 
the fecundity of adult and survival/ maturation of eggs simultaneously 
and thereby can increase the extinction risk of the population. For non-linear age-structured 
models, we got the similar trend for extinction risks with different tail association (left and right).
Non-linear model showed higher extinction risk than its corresponding 
linear analog as it was evident from Eq. \ref{eq_vitalrates} with finite $\Omega = 100$. 
Fig. \ref{SM-fig_extrisk_agestr_mtd} represents this scenario with two copulas having different 
asymmetric tails but same Kendall's correlation ($\tau = 0.6$). Extinction risks for
two different tail-dependent vital rates were not strinkingly different (e.g. see pair of red and
blue lines in Fig. \ref{SM-fig_extrisk_agestr_mtd}A or B) and we anticipated the difference being partly 
dependent on the degree of tail association. Therefore we repeated the same analysis with 
vital rates coming from extreme-tail dependent (left and right in two different simulations)
copulas and got the similar results but with much prominent difference between their extinction risks
(Fig. \ref{fig_extrisk_agestr_etd}). This result answered (Q1) and showed extinction risk of 
a population could be considered as direct manifestation of tail association property 
between vital rates of different age-class
of that population. <!--discussion later: add Tuljapurkar's approximation-->

@@

We found density-dependent metapopulation models in stochastic 
environment showed similar patterns for extinction risk as density-independent 
metapopulation model showed in recent study [@Ghosh_copula]. For each density-dependent 
model listed in Table \ref{tab_model_summary}, we observed higher extinction risk for 
the metapopulations when spatially correlated noises were left-tail dependent than the one with right-tail association.
But that observation remained valid when the model parameter - here considered as growth rate $(r)$ - was 
sufficiently low so that under-compensatory dynamics (monotonous approach towards
its stable equilibrium) for the deterministic analog of the model was ensured.
We observed right tail dependent noise caused higher extinction 
risk for the metapopulations than left tail dependent noise when the model 
was in over-compensatory regime (oscillatory approach towards stable equilibrium) for higher 
value of $r$. Figs. \ref{fig_extrisk_allmodel_localdisp} shows
variation of extinction risk against local dispersal rate ($d$) for six density dependent metapopulation models 
(see Table \ref{tab_model_summary}) with growth rate $r$ considered either from under-compensatory or 
from over-compensatory regime. We also got similar results with our prediction (see Q2 from \nameref{Introduction})
for global dispersal (Fig. \ref{SM-fig_extrisk_allmodel_globaldisp}). 
Figs. (\ref{SM-fig_ricker_risk_local_disp_scl_1_etd} - \ref{SM-fig_risk_local_disp_abrewer_r_vary_scl_0.6_mtd}) 
demonstrate how extinction risk changed against local dispersal rate for a whole range
of $r$, whereas Figs. (\ref{SM-fig_ricker_risk_global_disp_scl_1_etd} - 
\ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6_mtd}) 
represent the same but for global dispersal among patches. 
Both case showed the signature of "rescue-effect" (i.e. reduced extinction risk for 
finite dispersal compared with $d=0$).

<!--Q3 : metapopulation results : switching near rc for d=0-->
Following Figs. (\ref{SM-fig_ricker_risk_local_disp_scl_1_etd} - 
\ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6_mtd}) we hypothesized that the switching 
between the extinction risk generated by left-tail dependent and right-tail dependent noise 
occurred near the critical parameter $r_c$ (see Q3). For simplicity we only considered 
$d=0$ case, i.e., when there was no dispersal between patches (see \nameref{Methods}). 
Figs. \ref{fig_extrisk_d0_etd} and \ref{SM-fig_extrisk_d0_mtd} demonstrate higher extinction risk 
with left-tail dependent noise for $r<r_c$ and with right-tail dependent noise 
for $r>r_c$. $r_c$ was the vertical line in Figs. \ref{fig_extrisk_d0_etd} and \ref{SM-fig_extrisk_d0_mtd} 
indicating the value of $r$ for each model where their deterministic analog switches 
from under- to over-compensatory dynamics. Though the switching visually appered far away
from the theoretical $r_c$ value for Verhulst and Austin-Brewer model, they were 
consistent with our hypothesis as they had comperatively 
smaller range of $r$ than the other models.

@@

<!--Tuljapurkar approximation : linear version of age-structured model in stochastic environments-->
We found left tail association in vital rates (induced by ECEs) caused higher extinction risk than 
the right tail dependent ones, but the difference was not remarkable (Fig. \ref{SM-fig_extrisk_agestr_mtd}) 
unless the tail association was at its extreme (Fig. \ref{fig_extrisk_agestr_etd}). This finding could be explained 
by the approximation @tuljapurkar1990lecture made for the stochastic log(growth rate) in an iid environment.
The approximation considered growth rate being dependent on the most dominant eigen value of the mean projection matrix 
and the temporal variance of the log(population growth rate) caused by environmental stochasticity - which was based
on the linear covariance measure between any two matrix elements, not on their tail association properties. Therefore,
following @tuljapurkar1990lecture, population growth rate in an iid environment could be approximated
as same for our model (Eq. \ref{eq_age_str}) with left-tail or right-tail association.

@@

 <!--FIG : age structure model results : linear, nonlinear-->
\begin{figure*}[!h]
\textbf{ \hspace{5.4 cm} (A) \hspace{4 cm} (B)} \\
\vspace{-0.9 cm}
\begin{center}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_extremetail_linear.pdf}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_extremetail_nonlinear_omg_100.pdf}\\
\end{center}
\textbf{ \hspace{5.4 cm} (C) \hspace{4 cm} (D)} \\
\vspace{-0.9 cm}
\begin{center}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_C_SC_linear.pdf}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_C_SC_nonlinear_omg_100.pdf}
\caption{Extinction risk against time from single patch age structured linear ($\Omega = \infty$) $(A, C)$ and nonlinear ($\Omega = 100$) $(C, D)$ models for $10000$ simulations; shading represents 95\% confidence interval. $(A, B)$ show extinction risk with time when fecundity of adults and survival of eggs have extreme left tail (LT) or extreme right tail (RT) association, $(C, D)$ show the analogous scenario but with moderate tail association between vital rates. \label{fig_extrisk_agestr}}
\end{center}
\end{figure*}

@@

<!--FIG : all nonlinear model results, d=0, extrisk vs. r-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=12 cm]{./Results/common_legend_extreme_and_moderate_noise.pdf}\\
\includegraphics[width=6 cm]{./Results/ricker_scl_1/ricker_extrisk_d0_extreme_and_moderate_noise.pdf}
\includegraphics[width=6 cm]{./Results/hassell_scl_1/hassell_extrisk_d0_extreme_and_moderate_noise.pdf}\\
\vspace{-0.6 cm}
\includegraphics[width=6 cm]{./Results/msmith_scl_0.8/msmith_extrisk_d0_extreme_and_moderate_noise.pdf}
\includegraphics[width=6 cm]{./Results/pennycuick_scl_1/pennycuick_extrisk_d0_extreme_and_moderate_noise.pdf}\\
\vspace{-0.6 cm}
\includegraphics[width=6 cm]{./Results/verhulst_scl_0.6/verhulst_extrisk_d0_extreme_and_moderate_noise.pdf}
\includegraphics[width=6 cm]{./Results/austinbrewer_scl_0.6/abrewer_extrisk_d0_extreme_and_moderate_noise.pdf}\\
\caption{Transition of higher to lower extinction risk (calculated after $25$ years) for 
the metapopulations affected by extreme left tail dependent noise across
$5$ multiple patches without dispersal ($d=0$), as intrinsic 
growth rate ($r$) goes from under to over-compensatory regime. Extinction risk for 
extreme right tail dependent noise follows the opposite trend.
Error bars represent the standard errors associated with $10000$
simulations for extinction risk computed at each $r$. Dotted line represents the value of $r$
for each model where their deterministic analog switches from under- to over-compensatory dynamics.
\label{fig_extrisk_d0_etd}}
\end{center}
\end{figure*}

>>>>

<!--test hypothesis from Q2 : ext risk with density-dependence metapopulation model with r from different regime-->
To answer (Q2) of \nameref{Introduction} for each metapopulation model,
we estimated extinction risk for a wide range of parameter $r$ 
that covered the undercompensatory and overcompensatory regime. We 
chose equispaced values of $r$ from the 95\% quantiles 
of possible range ($r_{min},r_{bf}$), where $r_{min}$ was the minimum value of $r$ 
to get a positive, finite equilibrium $P_e$ (for details, see Appendix 
\ref{SM-model_stability}) to start with (at $t=0$). For each chosen $r$,
we calculated extinction risk after $25$ years with each metapopulation model 
where population could disperse locally (or globally) with varying dispersal 
rate $d= 0,0.1,0.2,..0.5$ and moderate to extreme left 
tail- and right-tail associated environmental noises with same Kendall's $\tau$ (= `r ktau_common`)
in different simulations (for code see [`mydispersal.R`](https://github.com/sghosh89/ERC/blob/master/mydispersal.R)).

We plotted extinction risks along y-axis
for both left tail and right tail associated noises 
against the varying range of $r$ along x-axis when no dispersal ($d=0$) was considered.
We excluded the possibility of dispersal to avoid any rescue effect among patches. 
For each model with extreme (or moderate) left- and right-tail associated noises, we calculated 
2.5$^{th}$ and 97.5$^{th}$ quantiles
of 10000 simulations with the probability of computed
extinction risks at each $r$. We compared $r_c$ and the value of $r$ at which 
95\% CI of extinction risks from both noises overlapped
with each other. This could give us an idea of how extinction risks affected by the model parameter 
$r$ that controlled the underlying dynamics of the model, i.e., undercompensatory 
or overcompensatory.

@@

For single patch linear age-structured models, we found higher extinction risk for moderate left-tail 
association (modeled by Clayton copula) in fluctuating vital 
rates compared with the one when vital rates were right tail dependent (modeled by Survival Clayton copula).
It was expected because an extreme "bad" year (e.g. with severe cold) can suppress 
the fecundity of adult and survival/ maturation of eggs simultaneously 
and thereby can increase the extinction risk of the population. For non-linear age-structured 
models, we got the similar trend for extinction risks with different tail association (left and right).
Non-linear model showed higher extinction risk than its corresponding 
linear analog as it was evident from Eq. \ref{eq_vitalrates} with finite $\Omega = 100$. 
Fig. \ref{fig_extrisk_agestr} represents this scenario when vital rates have moderate to extreme tail association in either tail but with same Kendall's correlation ($\tau =$ `r ktau_common`). Extinction risks for two different tail-dependent vital rates were not strinkingly different (e.g. see pair of red and
blue lines in Fig. \ref{fig_extrisk_agestr}A or B) and we anticipated the difference being partly 
dependent on the degree of tail association. Therefore we repeated the same analysis with 
vital rates coming from extreme-tail dependent (left and right in two different simulations)
copulas and got the similar results but with much prominent difference between their extinction risks
(Fig. \ref{fig_extrisk_agestr}C or D). This result answered (Q1) and showed extinction risk of 
a population could be considered as direct manifestation of tail association property 
between vital rates of different age-class
of that population. <!--discussion later: add Tuljapurkar's approximation-->

@@

We found density-dependent metapopulation models in stochastic 
environment showed similar patterns for extinction risk as density-independent 
metapopulation model did in recent study [@Ghosh_copula]. For each density-dependent 
model listed in Table \ref{tab_model_summary}, we observed higher extinction risk for 
the metapopulations when spatially correlated noises were left-tail associated 
than the one with right-tail association. But that observation remained 
valid when the model parameter - here considered as growth rate $(r)$ - was 
sufficiently low so that under-compensatory dynamics (monotonous approach towards
its stable equilibrium) for the deterministic analog of the model was ensured.
We observed right tail associated noise caused higher extinction 
risk for the metapopulations than left tail associated noise when the model 
was in over-compensatory regime (oscillatory approach towards stable equilibrium) for higher 
value of $r$. Fig. \ref{fig_extrisk_allmodel_localdisp} shows
variation of extinction risk against local dispersal rate ($d$) for 
six density dependent metapopulation models (see Table \ref{tab_model_summary}) with growth 
rate $r$ considered either from under-compensatory or 
from over-compensatory regime. This finding addresses (Q2) from \nameref{Introduction}. We 
also got similar results 
for global dispersal (Fig. \ref{SM-fig_extrisk_allmodel_globaldisp}). 
Figs. (\ref{SM-fig_ricker_risk_local_disp_scl_1} - 
\ref{SM-fig_risk_local_disp_abrewer_r_vary_scl_0.6}) 
demonstrate how extinction risk changed against local dispersal rate for a whole range
of $r$, whereas Figs. (\ref{SM-fig_ricker_risk_global_disp_scl_1} - 
\ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6}) 
represent the same but for global dispersal among patches. 
Both case showed the signature of "rescue-effect" (i.e. reduced extinction risk for 
finite dispersal compared with $d=0$).

<!--Q3 : metapopulation results : switching near rc for d=0-->
Following Figs. (\ref{SM-fig_ricker_risk_local_disp_scl_1} - 
\ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6}) one can reasonably hypothesize
that the switching between the extinction risk 
generated by left-tail associated and right-tail associated noise 
occurred near the critical parameter $r_c$. For simplicity, we only considered 
$d=0$ case, i.e., when there was no dispersal between patches. 
Fig. \ref{fig_extrisk_d0} demonstrates higher extinction risk 
with left-tail associated noise for $r<r_c$ and with right-tail associated noise 
for $r>r_c$. $r_c$ was the vertical line in Fig. \ref{fig_extrisk_d0} 
indicating the value of $r$ for each model where their deterministic analog switches 
from under- to over-compensatory dynamics. Though the switching visually appered far away
from the theoretical $r_c$ value for Verhulst and Austin-Brewer model, they were 
consistent with our hypothesis as they had comperatively 
smaller range of $r$ than the other models.

@@

<!--Tuljapurkar approximation : linear version of age-structured model in stochastic environments-->
We found left tail association in vital rates (induced by ECEs) caused higher extinction risk than 
the right tail dependent ones, but the difference was not remarkable 
unless the tail association was at its 
extreme (Fig. \ref{fig_extrisk_agestr}). This finding could be explained 
by the approximation @tuljapurkar1990lecture made for the 
stochastic log(growth rate) in an iid environment.
The approximation considered growth rate being 
dependent on the most dominant eigen value of the mean projection matrix 
and the temporal variance of the log(population growth rate) caused 
by environmental stochasticity - which was based
on the linear covariance measure between any 
two matrix elements, not on their tail association properties. Therefore,
following @tuljapurkar1990lecture, population growth 
rate in an iid environment could be approximated
as same for our model (Eq. \ref{eq_age_str}) with 
left-tail or right-tail association.

<!--Intution/ possible explanation for expected results/ hypo?-->

@@

 <!--FIG : age structure model results : linear, nonlinear-->
\begin{figure*}[!h]
\textbf{ \hspace{5.4 cm} (A) \hspace{4 cm} (B)} \\
\vspace{-0.9 cm}
\begin{center}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_C_SC_linear.pdf}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_C_SC_nonlinear_omg_100.pdf}\\
\end{center}
\textbf{ \hspace{5.4 cm} (C) \hspace{4 cm} (D)} \\
\vspace{-0.9 cm}
\begin{center}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_extremetail_linear.pdf}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_extremetail_nonlinear_omg_100.pdf}
\caption{Extinction risk against time from single patch age structured linear ($\Omega = \infty$) $(A, C)$ and nonlinear ($\Omega = 100$) $(B, D)$ models for $10000$ simulations; shading represents 95\% confidence interval. $(A, B)$ show extinction risk with time when fecundity of adults and survival of eggs have moderate left tail (LT) or moderate right tail (RT) association, $(C, D)$ show the analogous scenario but with extreme tail association between vital rates. \label{fig_extrisk_agestr}}
\end{center}
\end{figure*}

@@

<!--FIG : all nonlinear model results, d=0, extrisk vs. r-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=12 cm]{./Results/common_legend_extreme_and_moderate_noise.pdf}\\
\includegraphics[width=6 cm]{./Results/ricker_scl_1/ricker_extrisk_d0_extreme_and_moderate_noise.pdf}
\includegraphics[width=6 cm]{./Results/hassell_scl_1/hassell_extrisk_d0_extreme_and_moderate_noise.pdf}\\
\vspace{-0.6 cm}
\includegraphics[width=6 cm]{./Results/msmith_scl_0.8/msmith_extrisk_d0_extreme_and_moderate_noise.pdf}
\includegraphics[width=6 cm]{./Results/pennycuick_scl_1/pennycuick_extrisk_d0_extreme_and_moderate_noise.pdf}\\
\vspace{-0.6 cm}
\includegraphics[width=6 cm]{./Results/verhulst_scl_0.6/verhulst_extrisk_d0_extreme_and_moderate_noise.pdf}
\includegraphics[width=6 cm]{./Results/austinbrewer_scl_0.6/abrewer_extrisk_d0_extreme_and_moderate_noise.pdf}\\
\caption{Transition of higher to lower extinction risk (calculated after $25$ years) for 
the metapopulations affected by extreme left tail dependent noise across
$5$ multiple patches without dispersal ($d=0$), as intrinsic 
growth rate ($r$) goes from under to over-compensatory regime. Extinction risk for 
extreme right tail dependent noise follows the opposite trend.
Error bars represent the standard errors associated with $10000$
simulations for extinction risk computed at each $r$. Dotted line represents the value of $r$
for each model where their deterministic analog switches from under- to over-compensatory dynamics.
\label{fig_extrisk_d0}}
\end{center}
\end{figure*}
